Reachability analysis for continuous one-counter automata

### 2021-05-30 Documentation/paper.tex
finalised acceleration in both documentation and code + tests

\subsubsection{Interval updates}
From the initial state, we will start doing cycles in which we update all intervals once, wherever possible. This will be done using the successor function \textit{succ} below where $R_i: Q -> 2^\mathbb{Q}$ represent the mapping function for the i-th iteration. This function is going from each state to its corresponding counter reachability interval. The \textit{succ} function will update the interval of one state, \textit{q}, to its next iteration, based upon the current configuration.

\begin{gather*}
	\hspace{-190 pt} succ_q(R_i) := R_i(q)\\
	\hspace{60 pt} \cup \bigcup \{(R_i(r) + (0, z]) \cap \tau(q)\,|\,(r,z,q) \in T, z>0\}\\
	\hspace{60 pt} \cup \bigcup \{(R_i(r) + (z, 0]) \cap \tau(q)\,|\,(r,z,q) \in T, z<0\}\\
	\hspace{-20 pt} \cup \bigcup \{R_i(r) \cap \tau(q)\,|\,(r,0,q) \in T\}
\end{gather*}

This formula allows us to iteratively compute the next interval configuration to fully explore which states have a non empty interval. Consider the example given in Figure \ref{fig:example_succ}. We will initialise $R_0$ to have all empty intervals, other than the first which we will initialise to $[0, 0]$. Note that these intervals are stored per pair of final and preceding node and not, as the formula would suggest, per node in general. This is because this allows us to have a better tracking of which path has which effect, which will help us in later steps.

@@

		\node[round] 	at (0, 0) 		(q0) 	{$q_0$};
		\node[round] 	at (3, 0)		(q1) 	{$q_1$};
		\node[round] 	at (6, 0)		(q2) 	{$q_2$};
		\node[round] 	at (9, 0)		(q3) 	{$q_3$};
		\node[invis]	at (10, 0.5)		(q4)	{$\geq 100$};

@@

\begin{table}[h]
	\begin{tabular}{ |c|c|c|c|c|c|c|c| }
		\hline
		q		& r 	& $R_0$ 		& $R_1$			& $R_2$			& $R_3$			& $R_4$ 		& $R_5$ \\
		\hline
		$q_0$	& $q_0$ & [0, 0] 		& [0, 0] 		& [0, 0] 		& [0, 0]		& [0, 0]		& [0, 0] \\
		$q_1$	& $q_0$ & $\emptyset$ 	& (0, 1]		& (0, 1] 		& (0, 1]		& (0, 1] 		& (0, 1] \\
		$q_1$	& $q_2$ & $\emptyset$ 	& $\emptyset$ 	& $\emptyset$	& (0, 3]		& (0, 3] 		& (0, 5] \\
		$q_2$	& $q_1$ & $\emptyset$ 	& $\emptyset$ 	& (0, 2] 		& (0, 2]		& (0, 4] 		& (0, 4] \\
		$q_3$	& $q_2$ & $\emptyset$ 	& $\emptyset$ 	& $\emptyset$	& $\emptyset$	& $\emptyset$	& $\emptyset$ \\
		\hline
	\end{tabular}
	\centering
	\caption{Decision table for the scenario in which a guest occupies a new cell.}
	\label{table:table_example_succ}
\end{table}

For the first successor, we will only update for the transition going from $q_0$ to $q_1$ as this is the only transition where there is a preceding node that is not empty. All other transitions will both have an empty interval for the start and ends nodes which will therefore trivially result in empty intervals after applying the succ function. When we apply the succ function on the first interval, we will get $(0, 1]$ as a resulting interval. The 0 is not included as the $(0,z]$ with $z=1$ which we will add to it does not include 0. As $q_1$ has no condition the $\tau(q_1)$ will simply be $(-\infty, \infty)$ and will therefore have no effect.

The second successor will update for the transitions going from $q_1$ to $q_2$. This update will be similar to the update that happened for $q_0$ to $q_1$ in every way but the starting interval, which is now (0, 1] and will thus result in an interval of (0, 2].

@@

As can be seen in Table \ref{table:table_example_succ} a few more successors were computed and we can clearly notice that we will eventually reach an interval which will exceed the 100 bound which is placed upon $q_3$. Reaching this interval will however take us quite some time, and considering that this is only a small example, this is far from desired. Furthermore, this loop is not bounded, this means that it can infinitely expand. A solution for this is loop acceleration in which we will accelerate our loop based upon the behaviour said loop shows. This solution will be discussed within the next sections.

@@

To perform loop acceleration we need to consider the notion of \textit{expanding cycles}. A cycle or loop is considered to be \textit{positively expanding} if there is a state sequence $q_i, q_{i+1}, q_{i+2}, ..., q_n$ such that:
\begin{enumerate}
	\item $q_i = q_n$
	\item $(q_j, z_j, q_{j+1}) \in T$ for all $j \in [i, n-1]$
	\item $R_n \subseteq R_i$
	\item $sup(R_i) \geq sup(R_n)$
\end{enumerate}

Similarly we say that a cycle is \textit{negatively expanding} if the following hold:
\begin{enumerate}
	\item $q_i = q_n$
	\item $(q_j, z_j, q_{j+1}) \in T$ for all $j \in [i, n-1]$
	\item $R_n \subseteq R_i$
	\item $inf(R_i) \leq inf(R_n)$
\end{enumerate}

Given that we know whether or not a cycle is either positively or negatively expanding, we can look for the strongest bound relative to the direction in which the loop expands.

\subsubsection{Efficiency}

>>>>

\subsubsection{Interval updates}
\label{interval updates}
From the initial state, we will start doing cycles in which we update all intervals once, wherever possible. This will be done using the successor function \textit{succ} below where $R_i: Q \times Q \rightarrow 2^\mathbb{Q}$ represent the mapping function for the i-th iteration. This function is going from each transition to its corresponding counter reachability interval. The \textit{succ} function will update the interval of one transition, \textit{(p, z, q)}, to its next iteration, based upon the current configuration.

\begin{gather*}
	\hspace{-134 pt} succ_{p, q}(R_i) := \cup \bigcup \{R_i(q, l)\,|\,(q,z,l) \in T\}\\
	\hspace{60 pt} \cup \bigcup \{(R_i(p, q) + (0, z]) \cap \tau(q)\,|\,(p,z,q) \in T, z>0\}\\
	\hspace{60 pt} \cup \bigcup \{(R_i(p, q) + (z, 0]) \cap \tau(q)\,|\,(p,z,q) \in T, z<0\}\\
	\hspace{-20 pt} \cup \bigcup \{R_i(p, q) \cap \tau(q)\,|\,(p,0,q) \in T\}
\end{gather*}

This formula allows us to iteratively compute the next interval configuration to fully explore which states have a non empty interval. Consider the example given in Figure \ref{fig:example_succ}. We will initialise $R_0$ to have all empty intervals, other than the first which we will initialise to $[0, 0]$. Note that these intervals are stored per pair of final and preceding node and not per node in general. This is because this allows us to have a better tracking of which path has which effect, which will help us in later steps. In the end it suffices to know that a node is reachable, independent on the path via which it can be reached.

@@

		\node[round] 	at (0, 0) 		(q0) 	{$q_0$};
		\node[round] 	at (3, 0)		(q1) 	{$q_1$};
		\node[round] 	at (6, 0)		(q2) 	{$q_2$};
		\node[invis]	at (7, 0.5)		(q4)	{$\leq 200$};
		\node[round] 	at (9, 0)		(q3) 	{$q_3$};
		\node[invis]	at (10, 0.5)	(q4)	{$\geq 100$};

@@

\begin{table}[h]
	\begin{tabular}{ |c|c|c|c|c|c|c|c| }
		\hline
		p		& q 	& $R_0$ 		& $R_1$			& $R_2$			& $R_3$			& $R_4$ 		& $R_5$ \\
		\hline
		$q_0$	& $q_0$ & [0, 0] 		& [0, 0] 		& [0, 0] 		& [0, 0]		& [0, 0]		& [0, 0] \\
		$q_0$	& $q_1$ & $\emptyset$ 	& (0, 1]		& (0, 1] 		& (0, 1]		& (0, 1] 		& (0, 1] \\
		$q_1$	& $q_2$ & $\emptyset$ 	& $\emptyset$ 	& (0, 2] 		& (0, 2]		& (0, 4] 		& (0, 4] \\
		$q_2$	& $q_1$ & $\emptyset$ 	& $\emptyset$ 	& $\emptyset$	& (0, 3]		& (0, 3] 		& (0, 5] \\
		$q_2$	& $q_3$ & $\emptyset$ 	& $\emptyset$ 	& $\emptyset$	& $\emptyset$	& $\emptyset$	& $\emptyset$ \\
		\hline
	\end{tabular}
	\centering
	\caption{Decision table for the scenario in which a guest occupies a new cell.}
	\label{table:table_example_succ}
\end{table}

For the first successor, we will only update for the transition going from $q_0$ to $q_1$ as this is the only transition where there is a non empty predecessor. All other transitions will both have an empty interval for the start and ends nodes which will therefore trivially result in empty intervals after applying the succ function. When we apply the succ function on the first interval, we will get $(0, 1]$ as a resulting interval. The 0 is not included as the $(0,z]$ with $z=1$ which we will add to it does not include 0. As $q_1$ has no condition the $\tau(q_1)$ will simply be $(-\infty, \infty)$ and will therefore have no effect.

The second successor will update for the transition going from $q_1$ to $q_2$. This update will be similar to the update that happened for $q_0$ to $q_1$ in every way but the starting interval, which is now (0, 1] and will thus result in an interval of (0, 2].

@@

As can be seen in Table \ref{table:table_example_succ} a few more successors were computed and we can clearly notice that we will eventually reach an interval which will exceed the lower bound of 100 which is placed upon $q_3$. Reaching this interval will however take us quite some time, and considering that this is only a small example, this is far from desired. Furthermore, given a loop that is not bounded, it is possible that a loop has no fix point at all, which means that our algorithm will evaluate indefinitely. A solution for this is loop acceleration in which we will accelerate our loop based upon the behaviour said loop shows.

@@

To perform loop acceleration we need to consider the notion of \textit{expanding cycles}. A cycle or loop is considered to be \textit{positively expanding} if there is a path that follows a state sequence $q_i, q_{i+1}, q_{i+2}, ..., q_{n-1}, q_n$ evaluated in reachability iterations $R_0, R_{1}, R_{2}, ..., R_{k-1}, R_{k}$, where k represents the current iteration, such that:
\begin{enumerate}
	\item $q_i = q_n$
	\item $\forall j \in [i, n-1]: (p_j, z_j, q_{j+1}) \in T$
	\item $\forall j \in [i, n-1]: R_k(p_j, q_{j+1}) \ne \emptyset$
	\item $\exists (p, z, q) \in T: z > 0$
	\item $sup(R_k(q_{n-1}, q_{n})) > sup(R_{k-1}(q_{n-1}, q_n))$
\end{enumerate}

Similarly we say that a cycle is \textit{negatively expanding} if the following hold:
\begin{enumerate}
	\item $q_i = q_n$
	\item $\forall j \in [i, n-1] : (p_j, z_j, q_{j+1}) \in T$
	\item $\forall j \in [i, n-1]: R_k(p_j, q_{j+1}) \ne \emptyset$
	\item $\exists (p, z, q) \in T: z < 0$
	\item $inf(R_k(q_{n-1}, q_{n})) < inf(R_{k-1}(q_{n-1}, q_n))$
\end{enumerate}

The first condition in combination with the second condition simply implies that we require there to be an actual loop. The third condition states that we should have visited every transition at least once. In case we did not we have insufficient information needed to perform the acceleration and we will therefore only mark loops as expanding in case we have fully discovered it. This also eliminates the potential that we are looking at loops where some of the nodes are not reachable due to node constraints present within the loop. The fourth and fifth condition are there to ensure that there is an actual expansion. If only the fifth condition holds, it implies that there was a change in supremum due to something outside of the loop, in case only the fourth condition holds but not the fifth, it implies that we have already reached the bound of the loop. In the latter case the loop will not be marked as expanding as this implies that it reached its bound after one iteration.

Given that we know whether or not a cycle is either positively or negatively expanding, we can derive which node will reach its bound first. By doing so we can move the interval to this bound and thus greatly accelerating the evaluation of said loop. By performing one more iteration throughout the loop, all bounds would be reached and no further evaluation of the before mentioned loop will be needed.

To find the node who will be the first to reach a fix point interval, we can simply consider the node for which the current reachability interval's supremum/infimum is closest to the max/min node bound given that we are either trying to accelerate upwards or downwards respectively. In the formula below $reach(q)$ represents a function denoting the set of final counter values for a state $q$ and $\underline{\tau}(q_j)$ and $\overline{\tau}(q_j)$ represent the upper and lower node bounds of node $q_j$ respectively.\\
\\
\noindent
Given a loop that is expanding positively. If the following properties evaluate to true:

\begin{itemize}
	\item $sup(R_i) > sup(R_n)$
	\item $\exists (p, z, q) \in T: z > 0$
	\item $\overline{\tau}(q_k) - v_k = min_{i\leq j \le n}\,\overline{\tau}(q_j) - v_j$
\end{itemize}
The following property will apply:
\begin{itemize}
	\item $[v_k, \overline{\tau}(q_k)] \subseteq reach(q_k)$
\end{itemize}

\noindent
Given a loop that is expanding negatively. If the following properties evaluate to true:

\begin{itemize}
	\item $inf(R_i) < inf(R_n)$
	\item $\exists (p, z, q) \in T: z < 0$
	\item $v_k - \underline{\tau}(q_k) = min_{i\leq j \le n}\,v_j - \underline{\tau}(q_j)$
\end{itemize}
The following property will apply:
\begin{itemize}
	\item $[\underline{\tau}(q_k), v_k] \subseteq reach(q_k)$
\end{itemize}

In our approach this allows us to traverse the differences between the previous iteration and the current iteration whenever we update a node that we already have an interval for. Given the example that was explored in Section \nameref{interval updates} we can see that we have a loop going from $q_1$ to $q_2$ and back.

\begin{table}[h]
	\begin{tabular}{ |c|c|c|c|c|c|c|c| }
		\hline
		p		& q 	& $R_0$ 		& $R_1$			& $R_2$			& $R_3$			& $R_4$ 		& $R_5$ \\
		\hline
		$q_0$	& $q_0$ & [0, 0] 		& [0, 0] 		& [0, 0] 		& [0, 0]		& [0, 0]		& [0, 0] \\
		$q_0$	& $q_1$ & $\emptyset$ 	& (0, 1]		& (0, 1] 		& (0, 1]		& (0, 1] 		& (0, 1] \\
		$q_1$	& $q_2$ & $\emptyset$ 	& $\emptyset$ 	& (0, 2] 		& (0, 2]		& (0, 4] 		& (0, 4] \\
		$q_2$	& $q_1$ & $\emptyset$ 	& $\emptyset$ 	& $\emptyset$	& (0, 3]		& (0, 3] 		& (0, 5] \\
		$q_2$	& $q_3$ & $\emptyset$ 	& $\emptyset$ 	& $\emptyset$	& $\emptyset$	& $\emptyset$	& $\emptyset$ \\
		\hline
	\end{tabular}
	\centering
	\caption{Decision table for the scenario in which a guest occupies a new cell without acceleration.}
	\label{table:table_example_no_acc}
\end{table}

In Table \ref{table:table_example_no_acc} we can see that in the third reachability interval, $R_2$, a counter value interval is discovered for the transition going from $q_1$ to $q_2$. In the next reachability interval the same happens for $q_2$ to $q_1$. In $R_4$ we will once again update the interval for the transition going from $q_1$ to $q_2$ which will result in all conditions for a positive expanding loop evaluating to true. This is because we have a loop of which we know that the supremum of one of the transitions just increased and we have a transition part of this loop with a strictly positive operation.

Applying our previously described steps, we will find that node $q_2$ is closest to its bound and we will therefore choose to expand the interval of the transition ending in this node, being the transition from $q_1$ to $q_2$. The result is shown in Table \ref{table:table_example_acc}.

\begin{table}[t!]
	\begin{tabular}{ |c|c|c|c|c|c|c|c| }
		\hline
		p		& q 	& $R_0$ 		& $R_1$			& $R_2$			& $R_3$			& $R_4$ 		& $R_5$ \\
		\hline
		$q_0$	& $q_0$ & [0, 0] 		& [0, 0] 		& [0, 0] 		& [0, 0]		& [0, 0]		& [0, 0] \\
		$q_0$	& $q_1$ & $\emptyset$ 	& (0, 1]		& (0, 1] 		& (0, 1]		& (0, 1] 		& (0, 1] \\
		$q_1$	& $q_2$ & $\emptyset$ 	& $\emptyset$ 	& (0, 2] 		& (0, 2]		& (0, 200] 		& (0, 200] \\
		$q_2$	& $q_1$ & $\emptyset$ 	& $\emptyset$ 	& $\emptyset$	& (0, 3]		& (0, 3] 		& (0, 201] \\
		$q_2$	& $q_3$ & $\emptyset$ 	& $\emptyset$ 	& $\emptyset$	& $\emptyset$	& $\emptyset$	& [100, 200] \\
		\hline
	\end{tabular}
	\centering
	\caption{Decision table for the scenario in which a guest occupies a new cell with acceleration.}
	\label{table:table_example_acc}
\end{table}

If we do one additional iteration, we can see that we have now discovered an interval for the transition going from $q_2$ to $q_3$. We have furthermore achieved a fix point, as both $q_2$ and $q_3$ have met their bounds.

\subsubsection{Proof of termination}
This algorithm will always terminate. Given that there are no loops it will trivially terminate within n evaluations, given that n is equal to the total number of nodes within the automaton. If there are loops they must either be positively expanding, negatively expanding or not expanding at all. In the last case the simulation will trivially terminate as the loops will end in a fix point after their first evaluation. If they are expanding, we have shown that we are capable of accelerating them after which one additional evaluation will always make them reach their bounds.

\begin{figure}[b!]%
	\centering
	\begin{tikzpicture}[auto, >=latex, node distance = 1 cm]
		\tikzstyle{round} = [thick, draw=black, circle, font=\small]
		\tikzstyle{invis} = [draw=none, font=\small]
		\tikzstyle{arrow} = [thick,->,>=stealth]

		\node[round] 	at (0, 0)		(q1) 	{$q_0$};
		\node[round] 	at (-3, 0)		(q2) 	{$q_1$};
		\node[invis]	at (-3, 0.75)	(q3)	{$\leq 200$};
		\node[round] 	at (3, 1)		(q4) 	{$q_2$};
		\node[invis]	at (3, 1.5)		(q5)	{$\leq 200$};
		\node[round] 	at (3, -1)		(q6) 	{$q_3$};

		\path[->]
		(q1) 	edge 	[bend left]		node	[right, above] 	{$+2$} 	(q2)
		(q2)	edge 	[bend left]		node	[right, above]	{$+2$}	(q1)
		(q1) 	edge 	[bend left]		node 	[right, below]	{$-1$}	(q4)
		(q4) 	edge 	[bend left]		node	[right]		 	{$-1$} 	(q6)
		(q6)	edge 	[bend left]		node	[right, above]	{$-1$}	(q1);
	\end{tikzpicture}
	\caption{Example of an automaton in which post acceleration evaluation is needed.}
	\label{fig:example_recomputation_of_bounds}
\end{figure}

It is possible that further evaluations require bounds to be recomputed. For example, given the automaton described in Figure \ref{fig:example_recomputation_of_bounds}. In this automaton the loop between $q_0$ and $q_1$ will be expanded first. The second loop will still get evaluated and the lower bound of $q_0$ to $q_1$ will keep on decreasing, until the second loop will get accelerated, at which point a fix point will be achieved and the evaluation will end. This will always be the case given that we will always have a finite amount of nodes which will thus result in a finite amount of loops which will all accelerate at some point. The effects of said acceleration might need a few more evaluations to propagate their effect all across the automaton but once it does the evaluation will end. We can therefore conclude that the algorithm can never run indefinitely.

A mathematical proof for the the theoretical algorithm on which this approach is based is given in the paper by Michael Blondin et al. \cite{blondin2021continuous}. This paper will proof that this algorithm will always terminate, and will do so in at most $(4\vert Q\vert ² + 5\vert Q\vert)$ accelerations.
